Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-12-12 draft.tex
included Bernhards revision

% \author{Blind}
\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
\affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
\affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
\email{\{klemm, glasser\}@isg.cs.uni-magdeburg.de}
}

@@

\abstract{ 
Epidemiology aims to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
We assess potential relations between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as the spine canal course, its curvature and torsion.
%%
We analyze mutual dependencies among shape-describing variables and non-image variables, e.g., pain indicators.
%%
Therefore, we automatically extract a decision tree classifier for each non-image variable.
%%
We provide an IVA technique to compare classifiers with interactive scatter plot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient to predict lifestyle factors within the data.
%%
A correlation between lumbar spine shape and lower back pain could not be found with the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, relations between non-image variables were successfully detected and characterized.
}

@@

Therefore, statistical analysis has to take missing data into account.

@@

For example, continuous data, such as age, is often discretized into quantile bins for visualizing age-dependent disease frequencies.

@@

Epidemiologists want to characterize the relation between aging and degenerative process of the spine \cite{Szpalski2005}.

@@

\noindent \textbf{Visual Analysis of Medical Image and Non-Image data}

@@

\texttt{GPLOMs} provide an overview, but take up much screen space and are therefore only suitable for few variables at once (see example in Fig.~\ref{fig:image-parameter-range} later on).

@@

\noindent Epidemiologists compiled a data set with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects

@@

	\item 21 metric variables, describing somatometric variables and markers retrieved from blood analysis

@@

	\item 113 categorical variables divided into 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g., \emph{pancreatitis} or \emph{high blood pressure} and 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status

@@

For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables, allowing to determine which manifestation can be described best using the image-based variables

@@

	\item \emph{Mean Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{\textit{curvature}_i}{I}$. We refer to the total curvature as \emph{curvature}.

@@

	\item \emph{Mean Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{\textit{torsion}_i}{I}$. We refer to the total torsion as \emph{torsion}.

@@

	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line connecting middle point and top/bottom point as \emph{sides}.

@@

As first experiment we correlated the shape variable with the dichotomous back pain indicator using a \texttt{GPLOM}.

@@

The metric image-derived variables are pairwise visualized using scatter plots.

@@

The combination of the image variables with back pain is visualized as histogram at the left side of the matrix diagonal and as box plot on the right side.
Figure~\ref{fig:image-parameter-range} (right) shows the range of each variable as box plot.

@@

The method uses \emph{pearson product-moment} for two continuous variables, \emph{polyserial} correlation for one continuous and one categorical variable, and \emph{Polychoric} correlations for two categorical variables.

@@

Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure} and are omitted, since they are not statistically resilient.

@@

The optimal classification uses a few rules to precisely characterize the target variable.

@@

For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, we get a mean error of 10\%, even though it is very bad.

@@

@@

Many interdependent variables are sparse, such as \emph{medication of diabetes} or \emph{reason of early retirement}.

@@

Therefore, we provide a way to adjust the minimal number of subjects for each variable using a slider input.

@@

Furthermore, we map the number of subjects associated with a variable to point the diameter in the scatter plot.

@@

None of the pain indicators can be described reliably using the image-based variables.

@@

The only variable reliably classified in this group is \emph{gender}, which can be predicted with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.

@@

The four \emph{body size} groups could be characterized with an error of $0.48$, but the decision tree comprises 71 rules and imposes overfitting.

@@

Variables, such as subjects sized 139-164 cm, between 64 and 90 years of \emph{age} or \emph{nutrition}-related variables are dominantly populated by one manifestation.

@@

Classification using groups divided by \emph{gender} do not produce satisfying results.

@@

The starting age of smoking could be characterized well with an error between $0.25$ and $0.32$ for all \emph{BMI} groups except for subjects with a \emph{BMI} of \emph{$30.5-48$}.

@@

Using mostly \emph{mean curvature} and \emph{curvature angle}, the \emph{leg pain level} can be predicted using $14$ rules with an error of $0.46$ for obese subjects (\emph{BMI} higher than $30$).

@@

Subjects experiencing \emph{pain in the last seven days} can also be predicted for this group using the same variables with a tree consisting of $8$ rules and an error of $0.35$

@@

The analysis was performed for subjects groups of \emph{gender}, \emph{BMI}, and \emph{body size} to assess their influence on the lumbar spine shape.

The methods presented herein may be applied to comprehensive epidemiological data sets to learn more about mutual dependencies among variables and to generate hypotheses on potential associations and subgroups.

@@

The presented results indicate that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not sufficient to characterize lumbar back pain in the SHIP data set.

>>>>

% \author{Blind}
\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
\affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
\affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
\email{klemm@isg.cs.uni-magdeburg.de}
}

@@

\abstract{ 
Epidemiology aims to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
We assess potential relations between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as spine canal curvature and torsion.
%%
We analyze mutual dependencies among shape-describing variables and non-image variables, e.g., pain indicators.
%%
Therefore, we automatically extract a decision tree classifier for each non-image variable.
%%
We provide an IVA technique to compare classifiers with interactive scatter plot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient to describe lifestyle factors within the data.
%%
A correlation between lumbar spine shape and lower back pain could not be found with the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, relations between non-image variables were successfully detected and described.
}

@@

Statistical analysis has to take missing data into account.

@@

For example, continuous data, such as body height, is often discretized into quantile bins for visualizing age-dependent disease frequencies.

@@

Epidemiologists want to describe the relation between aging and degenerative process of the spine \cite{Szpalski2005}.

@@

\noindent \textbf{Visual Analysis of Medical Image and Non-Image Data.}

@@

\texttt{GPLOMs} provide an overview, but take up much screen space and are therefore only suitable for a few variables at once (see example in Fig.~\ref{fig:image-parameter-range}).

@@

\noindent Epidemiologists compiled a data set with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects.

@@

	\item 21 metric variables, describing somatometric variables and markers retrieved from blood analysis.

@@

	\item 113 categorical variables divided into 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g., \emph{pancreatitis} or \emph{high blood pressure} and 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status.

@@

For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables, allowing to determine which manifestation can be described best using the image-based variables.

@@

	\item \emph{Mean Curvature} is defined as the average curvature of all points describing the centerline: $\sum_{i=1}^I \frac{\textit{curvature}_i}{I}$. We refer to the mean curvature as \emph{curvature}.

@@

	\item \emph{Mean Torsion} (deviation of a curve from its current course) is defined as the average torsion of all points describing the centerline: $\sum_{i=1}^I \frac{\textit{torsion}_i}{I}$. We refer to the mean torsion as \emph{torsion}.

@@

	\item \emph{Curvature angle $\alpha$} is the angle defined by the middle point of the spine canal centerline as \emph{vertex} and the line connecting middle point and top/bottom point as \emph{sides}.

@@

As first experiment we correlated the shape variable with the dichotomous back pain indicator using a \texttt{GPLOM} (Fig.~\ref{fig:image-parameter-range}).

@@

The metric image-derived variables are pairwise visualized using scatter plots on the left side of the matrix diagonal.

@@

%The combination of the image variables with back pain is visualized as histogram at the left side of the matrix diagonal and as box plot on the right side.
The combination of the image variables with back pain is visualized as histogram in the last row and as box plot in the last column.

@@

The method uses \emph{Pearson product-moment} for two continuous variables, \emph{polyserial} correlation for one continuous and one categorical variable, and \emph{polychoric} correlations for two categorical variables.

@@

Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure} are omitted, since they are not statistically resilient.

@@

The optimal classification uses a few rules to precisely describe the target variable.

@@

For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, we get a mean error of 10\%, even though it is very imprecise.

@@

$N_m$ denotes the number of subjects in manifestation $m$.

@@

Many variables are sparse, such as \emph{medication of diabetes} or \emph{reason of early retirement}.

@@

Therefore, we provide a way to adjust the minimal number of subjects for each variable using a slider.

@@

Furthermore, we map the number of subjects associated with a variable to the diameter in the scatter plot.

@@

None of the pain indicators can be reliably described using the image-based variables.

@@

The only variable reliably classified in this group is \emph{gender}, which can be described with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.

@@

The four \emph{body size} groups could be described with an error of $0.48$, but the decision tree comprises 71 rules and imposes overfitting.

@@

Variables, such as subject size 139-164 cm, between 64 and 90 years of \emph{age} or \emph{nutrition}-related variables are dominantly populated by one manifestation.

@@

Classifications using groups divided by \emph{gender} do not produce satisfying results.

@@

The starting age of smoking could be described well with an error between $0.25$ and $0.32$ for all \emph{BMI} groups except for subjects with a \emph{BMI} of \emph{$30.5-48$}.

@@

Using mostly \emph{mean curvature} and \emph{curvature angle}, the \emph{leg pain level} can be described using $14$ rules with an error of $0.46$ for obese subjects (\emph{BMI} higher than $30$).

@@

Subjects experiencing \emph{pain in the last seven days} can also be described for this group using the same variables with a tree consisting of $8$ rules and an error of $0.35$

@@

The analysis was performed for subjects groups of \emph{gender}, \emph{BMI} and \emph{body size} to assess their influence on the lumbar spine shape.

The methods presented herein may be applied to comprehensive epidemiological data sets to investigate mutual dependencies among variables and to generate hypotheses on potential associations and subgroups.

@@

The presented results indicate that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not sufficient to describe lumbar back pain in the SHIP data set.
